---
title: "1. Biome layers and occurrence data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Biome layers and occurrence data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Cap BLAS / OpenMP threads to 1 for the vignette build.
# Reason: on Windows with R >= 4.6 the bundled OpenBLAS occasionally fails
# with "Memory allocation still failed after 10 retries" during vignette
# re-building when multiple threads try to allocate matrix memory in
# parallel (triggered here by the raster extracts behind biomes_classify()
# and the pairwise kappa loop in biomes_rank()). The error is
# non-deterministic, so the cap is applied in every vignette. Forcing a
# single thread avoids the contention without changing any numerical
# result.
Sys.setenv(OPENBLAS_NUM_THREADS = "1")
Sys.setenv(OMP_NUM_THREADS      = "1")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(biomes)

# Several chunks below need the ~36 MB biome raster, which is downloaded
# on first use. Render them only when the raster can be fetched without
# burdening CRAN: locally (NOT_CRAN = "true") we download it; on CRAN
# these chunks are skipped, so the vignette builds without network access.
run_raster <- isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
if (run_raster) {
  run_raster <- tryCatch(
    {
      biomes_download(quiet = TRUE)
      TRUE
    },
    error = function(e) FALSE
  )
}
```

# Goal

This is the first of three vignettes that walk through the `biomes`
package:

1. **Biome layers and occurrence data** (this vignette): the raw
   material: load the biome layers, read their metadata, get an
   occurrence dataset, and decide *which* biome definition fits your
   data best.
2. [Classify, summarize and map](classify-summarize-map.html): assign
   occurrences to biomes, tabulate them, and draw a map.
3. [The one-call workflow](one-call-workflow.html): do all of the above
   in a single `biomes_full()` call.

Here we cover everything you need *before* classifying: the 31 biome
layers, their metadata, where occurrences come from, and the ranking
that suggests the most informative layer for a dataset.

---

# 1. The 31 biome layers

`biomes_get()` returns the packaged raster stack: 31 biome definitions
(Fischer et al. 2022) at 10 × 10 km resolution, globally.

```{r, eval = run_raster}
layers <- biomes_get()
layers
```

It is an ordinary `terra::SpatRaster`, so you can subset and plot single
layers directly (`biomes` attaches **terra** for you):

```{r, eval = run_raster, fig.alt = "World map of biome layer 1", fig.width = 7, fig.height = 4}
layers[[1]]
plot(layers[[1]])
```

---

# 2. Layer metadata

Each layer of the stack matches one row of `biomes_information`, in the
same order. Use it to find out which publication and classification a
layer comes from:

```{r}
data(biomes_information)
biomes_information[1, c("publication", "name_of_classification",
                        "layer_in_raster_stack")]
```

`biomes_info()` prints a human-readable summary for one, several, or all
layers:

```{r}
biomes_info(1)
biomes_info(c(1, 14, 21))
```

The class-level lookup (raster value → biome name, per layer) lives in
`biomes_legend`. You rarely call it directly; `biomes_classify()` and
`biomes_visualise()` use it internally to label biomes, but it is there
if you need it:

```{r}
data(biomes_legend)
head(biomes_legend)
```

---

# 3. Occurrence data

Every downstream function works on a table of occurrence records with a
longitude and a latitude column. The package ships a small example set,
`biomes_example`, so you can try everything without a download:

```{r}
data(biomes_example)
nrow(biomes_example)
head(biomes_example)
```

## Fetching occurrences from GBIF (optional)

If you do not already have a dataset, `biomes_occ()` can pull one from
GBIF for a taxon (species, genus, family, …) and run basic coordinate
cleaning. It asks GBIF how many records exist and, for up to 100,000,
downloads them via `occ_search()` (no login). This is a convenience
add-on; the rest of the package works on *any* occurrence table.

```{r, eval = FALSE}
occ <- biomes_occ(taxon = "Fagus sylvatica")
```

The result has the same shape as `biomes_example` (a `family`, `genus`,
`species`, `countryCode`, `decimalLongitude`, `decimalLatitude` table by
default), so it drops straight into the next steps.

---

# 4. Which biome definition fits my data?

With 31 schemes to choose from, `biomes_rank()` scores them for *your*
occurrences and proposes a single best layer. It rates each layer on
coverage (share of records classified), the effective number of biome
classes, and granularity, then combines these into a composite score.

```{r, eval = run_raster}
r       <- biomes_rank(biomes_example, verbose = FALSE)
best_id <- attr(r, "best_layer")
best_id
head(r)
```

`biomes_show_rank()` visualizes the ranking from three angles:

```{r, eval = run_raster && requireNamespace("ggplot2", quietly = TRUE), fig.width = 7, fig.height = 4}
biomes_show_rank(r, type = "composite")   # composite score per layer
biomes_show_rank(r, type = "na")          # % unclassified per layer
biomes_show_rank(r, type = "criteria")    # heatmap of all criteria
```

The integer in `attr(r, "best_layer")` is exactly the `layer` index you
pass to the classification and mapping functions in the next vignette.

---

# Next

You now have (a) the biome layers and their metadata, (b) an occurrence
dataset, and (c) a recommended layer. Continue with
[Classify, summarize and map](classify-summarize-map.html) to turn this
into a biome table and a map.
