---
title: "2. Classify, summarize and map"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Classify, summarize and map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Cap BLAS / OpenMP threads to 1 for the vignette build.
# Reason: on Windows with R >= 4.6 the bundled OpenBLAS occasionally fails
# with "Memory allocation still failed after 10 retries" during vignette
# re-building when multiple threads try to allocate matrix memory in
# parallel (triggered here by the raster extracts behind biomes_classify()
# and the pairwise kappa loop in biomes_rank()). Forcing a single thread
# avoids the contention without changing any numerical result.
Sys.setenv(OPENBLAS_NUM_THREADS = "1")
Sys.setenv(OMP_NUM_THREADS      = "1")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(biomes)

# Several chunks below need the ~36 MB biome raster, which is downloaded
# on first use. Render them only when the raster can be fetched without
# burdening CRAN: locally (NOT_CRAN = "true") we download it; on CRAN
# these chunks are skipped, so the vignette builds without network access.
run_raster <- isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
if (run_raster) {
  run_raster <- tryCatch(
    {
      biomes_download(quiet = TRUE)
      TRUE
    },
    error = function(e) FALSE
  )
}
```

# Goal

This is the second of three vignettes. The
[first one](biome-data.html) loaded the biome layers, an occurrence
dataset, and used `biomes_rank()` to pick a layer. Here we use that
layer to:

- **classify** occurrences into biomes with `biomes_classify()`,
- **summarize** them per biome with `biomes_tab()`, and
- **map** them with `biomes_visualise()`.

Throughout we use the bundled `biomes_example` dataset and pick a layer
by its index (`1:31`), the same index `biomes_rank()` returns as the
best layer.

```{r}
data(biomes_example)
```

---

# 1. Classify occurrences into biomes

`biomes_classify()` takes a table of points (or an `sf` / `SpatVector`)
and returns the **input data with the biome assignment appended on the
right**. You pick the layer by index; no need to handle `SpatRaster`
objects yourself.

```{r, eval = run_raster}
class_1 <- biomes_classify(biomes_example, layer = 1)
head(class_1)
```

A new column `Biome_Inventory_layer_01_name` has been added. Records
that fall outside every biome polygon are labelled `"no_biome"` rather
than dropped, so the counts stay complete:

```{r, eval = run_raster}
table(class_1$Biome_Inventory_layer_01_name, useNA = "ifany")
```

## Common variations

```{r, eval = run_raster}
# Several layers at once, one column per layer
class_multi <- biomes_classify(biomes_example, layer = c(1, 25))
head(class_multi)

# Keep both the raster value and the name
biomes_classify(biomes_example, layer = 1, value = "both") |> head(3)

# Return only the classification columns (drop the input)
biomes_classify(biomes_example, layer = 1, append = FALSE) |> head(3)

# Keep NA for off-polygon points instead of the "no_biome" label
class_na <- biomes_classify(biomes_example, layer = 1, na = NA)
sum(is.na(class_na$Biome_Inventory_layer_01_name))
```

Appended columns are named after the input layer with the suffixes
`_value` (raster value) and `_name` (biome name). For custom rasters
outside the packaged stack, pass a `terra::SpatRaster` via `biome =`.

---

# 2. Tabulate occurrences per biome

`biomes_tab()` counts **occurrence records** (one input row = one
occurrence) per biome and layer, returning a long table with one row per
(layer, biome) pair:

```{r, eval = run_raster}
biomes_tab(class_1)
```

It handles multi-layer classifications in the same call:

```{r, eval = run_raster}
biomes_tab(class_multi)
```

Because off-polygon points are labelled `"no_biome"`, every record is
counted. To count **unique species** per biome instead of records,
deduplicate by `species` before tabulating:

```{r, eval = run_raster && requireNamespace("dplyr", quietly = TRUE)}
library(dplyr)

class_1 |>
  distinct(species, Biome_Inventory_layer_01_name) |>
  biomes_tab()
```

---

# 3. Map occurrences over a biome layer

`biomes_visualise()` draws the chosen biome layer as a categorical
raster, overlays the occurrence points, and appends the per-biome record
counts to the legend labels (so they match `biomes_tab()` exactly).

```{r, eval = run_raster && all(vapply(c("sf", "ggplot2", "viridis", "tidyterra"), requireNamespace, logical(1), quietly = TRUE)), fig.width = 7, fig.height = 4, fig.alt = "Occurrences mapped over biome layer 1"}
biomes_visualise(biomes_example, layer = 1)
```

By default the map shows the biome colours, the points, and a legend on
the right. The pie inset is **off by default** (`pie = FALSE`); switch it
on to show the share of records per biome (segments ≥ 5 % are labelled):

```{r, eval = run_raster && all(vapply(c("sf", "ggplot2", "viridis", "tidyterra", "cowplot", "ggforce"), requireNamespace, logical(1), quietly = TRUE)), fig.width = 7, fig.height = 4, fig.alt = "Biome map with pie inset"}
biomes_visualise(biomes_example, layer = 1, pie = TRUE)
```

Drop the legend for a clean publication figure:

```{r, eval = FALSE}
biomes_visualise(biomes_example, layer = 1, legend = FALSE)
```

The function returns a `ggplot` object (or a `cowplot` object when
`pie = TRUE`), so you save it the usual way:

```{r, eval = FALSE}
p <- biomes_visualise(biomes_example, layer = 1)
ggplot2::ggsave("map_layer_01.jpg", p, width = 13, height = 8, dpi = 600)
```

---

# Next

You have now run the full building-block workflow: classify → tabulate →
map. The [third vignette](one-call-workflow.html) shows how
`biomes_full()` chains exactly these steps (plus the ranking from
vignette 1) into a single call.
