---
title: "3. The one-call workflow with biomes_full()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. The one-call workflow with biomes_full()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
# Cap BLAS / OpenMP threads to 1 for the vignette build.
# Reason: on Windows with R >= 4.6 the bundled OpenBLAS occasionally fails
# with "Memory allocation still failed after 10 retries" during vignette
# re-building when multiple threads try to allocate matrix memory in
# parallel (triggered here by the raster extracts behind biomes_classify()
# and the pairwise kappa loop in biomes_rank()). The error is
# non-deterministic, so the cap is applied in every vignette. Forcing a
# single thread avoids the contention without changing any numerical
# result.
Sys.setenv(OPENBLAS_NUM_THREADS = "1")
Sys.setenv(OMP_NUM_THREADS      = "1")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(biomes)
```

# Goal

The first two vignettes ran the workflow step by step:

1. [Biome layers and occurrence data](biome-data.html): load layers,
   get occurrences, rank the layers.
2. [Classify, summarize and map](classify-summarize-map.html):
   classify, tabulate, visualize.

`biomes_full()` chains all of those steps into a **single call**. Give
it an occurrence dataset (or a taxon name) and it picks a layer,
classifies the records, tabulates them, and builds a map, returning
everything in one object.

---

# 1. From an occurrence dataset

The simplest entry point: pass a data frame as `x`. With the default
`layer = "best"`, `biomes_full()` runs `biomes_rank()` internally and
uses the top-ranked layer.

```{r, eval = FALSE}
data(biomes_example)

res <- biomes_full(x = biomes_example)   # layer = "best" (default)
res
```

```
#> <biomes_full result>
#>   occurrences : 29104 records
#>   layer       : 16
#>   picked by   : biomes_rank() (composite = 0.986)
#>   table rows  : 20 (biomes used)
#> Components: $occ, $layer, $ranking, $classified, $table, $map
```

The result is a `biomes_full` object, really just a list with named
components you can pull apart:

```{r, eval = FALSE}
res$layer       # which layer biomes_rank() picked
res$ranking     # the full ranking table (NULL when layer is fixed)
res$classified  # the occurrences with the biome column appended
res$table       # occurrences per biome on the chosen layer
res$occ         # the occurrence data that went in
res$map         # the ggplot / cowplot map
print(res$map)
```

To use a fixed layer instead of ranking, pass an integer in `1:31`:

```{r, eval = FALSE}
biomes_full(x = biomes_example, layer = 31)
```

`biomes_full()` returns its result **invisibly** and does not print
anything by default. Set `show = TRUE` to also print the map and table
as a side effect:

```{r, eval = FALSE}
biomes_full(x = biomes_example, layer = 1, show = TRUE)
```

> **Note:** the map produced by `biomes_full()` follows the
> `biomes_visualise()` default, so it has **no pie inset**. If you want
> one, build the map yourself with
> `biomes_visualise(..., pie = TRUE)` (see vignette 2).

---

# 2. From a taxon name

Instead of `x`, pass a scientific name as `taxon`. `biomes_full()` then
calls `biomes_occ()` to download and clean GBIF occurrences first, and
proceeds exactly as above. This needs a network connection (and the
`rgbif` / `CoordinateCleaner` packages), so it is not run here:

```{r, eval = FALSE}
res <- biomes_full(taxon = "Fagus sylvatica")
res$table
print(res$map)

# multiple taxa, fixed layer
biomes_full(taxon = c("Suricata suricatta", "Mustela lutreola"), layer = 1)
```

Extra arguments are forwarded to `biomes_occ()`, e.g. `year_min`,
`year_max`, `limit`, `use_download`, or GBIF credentials:

```{r, eval = FALSE}
biomes_full(taxon = "Fagus sylvatica", year_min = 2000)
```

Provide **either** `x` **or** `taxon`, never both.

---

# 3. How it maps onto the building blocks

`biomes_full(x = data, layer = "best")` is equivalent to:

```{r, eval = FALSE}
ranking <- biomes_rank(data, verbose = FALSE)        # vignette 1
best_id <- attr(ranking, "best_layer")

classified <- biomes_classify(data, layer = best_id) # vignette 2
tab        <- biomes_tab(classified)                 # vignette 2
map        <- biomes_visualise(data, layer = best_id)# vignette 2
```

Use `biomes_full()` when you want the standard pipeline with one call;
reach for the individual functions when you need to tweak a step, e.g. a
different ranking criterion, custom classification columns, or a map
with the pie inset or without the legend.
